Versioned data: why it is needed and how it can be achieved (easily and cheaply)

### 2017-11-09 ms.tex
edits to the discussion

The main issue we have identified in this article will not be new for many readers: datasets are constantly improving and, despite tremendous advances in data sharing and associated technologies over the last decade, there is little consensus about how to handle dataset updates. Some specific solutions for versioning do already exist, such as with genetic sequence data on \texttt{GenBank} (see \smurl{www.ncbi.nlm.nih.gov/genbank/sequenceids/}). But so far, versioning solutions have only been developed for large dynamic databases managed by well-funded institutions or consortia and involving simple standardised data types. While individual research groups working with a specialized type of data could in principle create their own dynamic web interface, the technological hurdles, cost and maintenance required (Table \ref{tab:publishing_models}) will be discouraging for most groups. This suggests there is a substantial need for an easy, cheap, and scalable solution for serving versioned data. By adopting best practices from software engineering, we believe a workable system now exists. Moreover, we created the \texttt{datastorr} package to make such a \textsc{vdd} system easy to implement for \texttt{R} users. As it builds off established and open source software and data science platforms, the proposed system is already easy to deploy on relatively large scale.

A central feature of the proposed  \textsc{vdd} system is that data are maintained in the cloud. This has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it allows for hosting of a stream of data releases for distribution (Fig. \ref{fig:technology_stack}). Cloud systems thus act as a central point for the collection, curation, and distribution of the data. Additionally, one of the greatest benefits of using cloud-based tools like  \texttt{GitHub} for development of software and data has been the way they encourage contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013, Perkel-2016}. Multiple users can make changes to different parts of the code (or in our case, data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved. Adopting a \textsc{vdd} system thus has the added benefit of facilitating seamless and transparent collaboration among research groups in the construction and maintenance of datasets.

An important concern for any data delivery system is the stability and reliability of the system. In the short term, users want minimal downtime, high speed, and seamless operation. As one of the largest companies hosting computer source code, \texttt{GitHub} provides exceptional performance in this regard -- certainly as good or better than nearly any system scientists might build themselves. In the long term, scientists desire long-term preservation of datasets, software, and published papers. To enhance long-term stability of data-versions released on \texttt{GitHub}, users can also choose to automatically archive every version in one of several traditional data archives, with a \textsc{doi} (Digital Object Identifier) minted for each release. Currently, both \texttt{Zenodo} (\smurl{zenodo.org}) and  \texttt{FigShare} \smurl{figshare.com} each integrate with \texttt{GitHub} for archiving of material hosted there.

The design choices we made in datastorr represent only one of many possible ways to adopt a \textsc{vdd} system; we can imagine many alternative models for interacting with versioned data stored in the cloud. The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data. Indeed, as with every technology the best available approach is certain to evolve, especially as merging technologies---such as CKAN, OKFN or Git-LFS---facilitate even better \textsc{vdd} systems in the future.  There is also rapid current work on new implementations which will make versioning systems faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently \citep{Fli, Dat}.

>>>>

The main issue we have identified in this article will not come as news to many readers: datasets are constantly improving and, despite tremendous advances in data sharing and associated technologies over the last decade, there is little consensus about how to handle dataset updates. Some specific solutions for versioning already exist, such as those designed for genetic sequence data (e.g., \texttt{GenBank}; see \smurl{www.ncbi.nlm.nih.gov/genbank/sequenceids/}). But so far, versioning solutions have only been developed for large dynamic databases managed by well-funded institutions or consortia and involving simple standardised data types. While individual research groups working with a specialized type of data could in principle create their own dynamic web interface, the technological hurdles, cost and maintenance required (Table \ref{tab:publishing_models}) will be discouraging for most. This suggests there is a substantial need for an easy, cheap, and scalable solution for serving versioned data. By adopting best practices from software engineering, we believe a workable system now exists. Moreover, we created the \texttt{datastorr} package to make such a \textsc{vdd} system easy to implement for \texttt{R} users. As it builds off established and open source software and data science platforms, the proposed system is already easy to deploy on relatively large scale.

A central feature of the proposed  \textsc{vdd} system is that data are maintained in the cloud. This has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it allows for hosting of a stream of data releases for distribution (Fig. \ref{fig:technology_stack}). Cloud systems thus act as a central point for the collection, curation, and distribution of the data. Additionally, one of the greatest benefits of using cloud-based tools like  \texttt{GitHub} for development of software and data has been the way they encourage contributions from multiple individuals working simultaneously --- including from people from outside the initial group of project participants \citep{Rogers-2013, Perkel-2016}. Multiple users can make changes to different parts of the code (or in our case, data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved. Adopting a \textsc{vdd} system thus has the added benefit of facilitating seamless and transparent collaboration among research groups in the construction and maintenance of datasets.

An important concern for any data delivery system is the stability and reliability of the system. In the short term, users want minimal downtime, high speed, and seamless operation. As one of the largest companies hosting computer source code, \texttt{GitHub} provides exceptional performance in this regard -- certainly as good or better than nearly any system scientists might build themselves. In the long term, scientists want their datasets, software, and papers to preserved and remain accessible. To enhance long-term stability of data-versions released on \texttt{GitHub}, users can also choose to automatically archive every version in one of several traditional data archives, with a \textsc{doi} (Digital Object Identifier) minted for each release. Currently, both \texttt{Zenodo} (\smurl{zenodo.org}) and  \texttt{FigShare} \smurl{figshare.com} each integrate with \texttt{GitHub} for archiving of material hosted there.

The design choices we made in datastorr represent only one of many possible ways to adopt a \textsc{vdd} system; we can imagine many alternative models for interacting with versioned data stored in the cloud. The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data. Indeed, as with every technology the best available approach is certain to evolve, especially as merging technologies --- such as CKAN, OKFN or Git-LFS --- facilitate even better \textsc{vdd} systems in the future.  There are some really exciting ideas that are currently under development which will make versioning systems faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently \citep[e.g.,][]{Fli, Dat}.
